Transport map accelerated adaptive importance sampling, and application to inverse problems arising from multiscale stochastic reaction networks

### 2019-01-17 TETAIS.tex
Final edits complete, including spell check

In this paper, we seek to develop an approach to
tackle this problem when using adaptive importance sampling methods,
by using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds. This approach is applicable to a whole range of
problems for which Monte Carlo Markoc chain (MCMC) methods mix slowly.

@@

One common approach is to induce the quasi-equillibrium approximation
QEA. This approximation makes the assumption that fast reactions
enter quasi-equilibrium on a timescale which is negligible with
respect to the timescale on which the slow dynamics in the system are
evolving. This amounts to taking the asymptotic limit that the
propensities of the ``slow'' reaction channels are equal to zero. This
allows us to sample approximate trajectories of the slow reactions
without the need to compute many fast reaction events.

@@

We consider several different algorithms for sampling from these
distributions. First we implement both the ETAIS and MH algorithms with
a Gaussian proposal distribution. In the case of the ETAIS algorithm,
this is a Gaussian mixture proposal distribution, and in the MH
algorithm this is equivalent to a vanilla random walk. The proposal
distribution uses a covariance matrix which has been constructed using
the sample covariance of a sample produced using a standard MH-RW
algorithm. This is necessary since these algorithms otherwise fail to
converge in a reasonable time for
this problem. We also compare the ETAIS and MH algorithms when using a
transport map proposal distributionm, as was discussed in detail in Chapter~\ref{sec:TETAIS}.

@@

The convergence analysis has been performed for the constrained
approximation of the posterior distribution. We now look at the
differences in the approximations of the posterior distribution
arising in this problem between the CMA and QEA. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
\[
	\hat{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{\alpha}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
\]
This difference in the denominator causes a shift in the manifold on
which the posterior is concentrated, as can be seen in
Figure~\ref{fig:chem_diff}. The figure shows the difference in
posteriors densities
\begin{equation}\label{eqn:chem_diff}\pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
\end{equation}

@@

Since the two posteriors have been approximated using an MCMC sample,
there is a significant amount of noise, particularly in the tails of
the distributions. We can see that the differences in the marginals
for $k_1, k_2,$ and $k_3$ are relatively small, but the marginal for
$k_4$ varies by a significant amount.

@@

Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals}
we see that we should of course prefer the model $\mathcal{M}_0$ in
which we observe all reactions perfectly. However this model is not
significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$,
~\cite{kass1995bayes}). The Bayes factor $B_{0,2} = 4.1 > 3.2$ tells
us that we should significantly prefer the full model to the QEA
approximation of the posterior. These Bayes factors present a
relatively weak
argument that the constrained approximation of the posterior is
superior to the QEA approximation.

@@

Figure~\ref{fig:chem_model_comp} displays more compelling evidence in
this direction. Here we show the marginal distributions for the
quantities $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$ with
respect to each of the three posterior distributions, $\pi_i(\cdot|D,
\mathcal{M}_i)$, $i = 0, 1, 2$.

Firstly, in
Figure~\ref{fig:chem_model_comp} (a), displaying the marginals for
$\hat{k}_4^{\text{QEA}}$ we see that $\pi_0$, the density arising from
the full model with all reactions observed, is peaked a distance away
from $\pi_2$, QEA approximation. The CMA approximation $\pi_1$, in contrast, is
peaked close to peak of $\pi_0$, but there is more uncertainty here
since we do not observe the fast reactions.

In contrast, in Figure~\ref{fig:chem_model_comp} (b), the marginal
densities for $\hat{k}_4^{\text{CMA}}$ for $\pi_0$ and $\pi_1$ are
barely distinguishable, even with the vastly reduced data available in
the posterior $\pi_1$ with CMA approximation. The QEA approximation
$\pi_2$ is again peaked far away from both of the other distributions.
This appears to demonstrate that the data regarding the slow variable
$S$ is only informative about $k_1 $ and $\hat{k}_4^{\text{CMA}}$, as
predicted by the CMA multiscale approximation.

\subsection{Gene Regulatory Network}\label{sec:grn}
In this example, we look at a model of a gene regulatory network (GRN)
mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering}
used by a particular cell to regulate the amount of a protein, $P$,
present. Proteins can bind in pairs to form a dimer $D$. The dimer can
bind to the gene $G$ to a ``switched off'' state $G'$, in which the
production of mRNA molecules $M$ is inhibited. The mRNA encodes for
production of the protein $P$, and both $P$ and $M$ can degrade. 
The eight reactions are given in Equation~\eqref{eqn:GRN_reactions}.
\begin{align}
	P+P \xleftrightarrows[k_2]{k_1} D \quad & \quad G+D \xleftrightarrows[k_4]{k_3} G' \nonumber\\
	G \xrightarrow{k_5} G+M \quad & \quad M\xrightarrow{k_6} M+P  \label{eqn:GRN_reactions} \\
	P \xrightarrow{k_7} \emptyset \quad & \quad M \xrightarrow{k_8} \emptyset \nonumber
\end{align}

We create trajectories of this system with the following parameters:
\begin{eqnarray*}
k_1 = 0.04, \qquad
k_2 = 5000, \qquad
k_3 = 100, \qquad
k_4 = 1, \\ 
k_5 = 0.5, \qquad
k_6 = 2, \qquad
k_7 = 0.2, \qquad
k_8 = 0.05,
\end{eqnarray*}
over the time period $t \in [0,500]$.

T-cells can be manufactured which bind to
proteins of a certain type, and cause them to phosphoresce, allowing
for approximate observations of the concentration levels of that
protein. However, the T-cells will not necessarily differentiate
between monomers and dimers of the protein. We replicate such a
scenario, in a zero-observational noise setting, by assuming that we
can only observe $T=P + 2D + 2G'$, the total number of protein
molecules in the system, alongside $M$, the concentration of mRNA. The
number of dimers, and the status of the genetic switch, are assumed
unobservable. In
order to sample from the posterior distribution on the reaction
parameters in the system given these observations, we can either integrate over all possible
trajectories of the fast variables (which is computationally
intractable), or as in Section \ref{sec:chem_multiscale} we can use a
multiscale method to approximate the posterior.

@@

\subsubsection{Target Distribution}
As alluded to in the previous section, we aim to sample from the CMA
approximation of the posterior distribution as defined in
Equation~\eqref{eqn:chem_posterior}. This example is more involved
than the previous one, since we have an eight dimensional parameter
space corresponding to the eight reaction rates in the system
\eqref{eqn:GRN_reactions}. However half of those reactions are
unobservable, and their effect on the posterior is only felt through
their effect on the effective propensities $\hat{\alpha}_5$ and $\hat{\alpha}_7$.

The priors for this problem were chosen to be fairly uninformative with respect to the likelihood function for each reaction rate. A list of the hyper-parameters corresponding to each Gamma prior can be found in Table~\ref{tab:grn_priors}.

@@

The one- and two-dimensional marginal distributions for the full
posterior distribution are displayed in
Figure~\ref{fig:GRN_posterior}. The posterior does not contain many
interesting correlation structures, however several dimensions have
leptokurtic tails which are difficult for the standard ETAIS algorithm
to sample from. The marginal densities also vary over very different
scales, which could lead to problems with the algorithms which do not
benefit from the transport map.

@@

The transport map setup has not changed from the previous example. However, due to the scaling of the weight we have had to be more careful about which samples we use to build our map. We require that the objective function $C(T)$, from Equation~\eqref{eqn:TETAIS_objective} is convex, which requires that the second derivative, given in Equation~\eqref{eqn:TETAIS_hessian}, is positive definite. This positive definite property depends on all weights in our sample being strictly positive, which is not always possible on a computer. For this reason we filter out any samples from the optimisation problem where the weight is a numeric zero. This does not affect the validity of the map since a weight zero would not contribute to the expectation.

@@

The first thing to note is that the first ETAIS measurement happens
after 2500 samples, whereas the first MH measurement occurs after 50
samples. This is due to the much larger ensemble size required to
sample in higher dimensions. We believe that an ensemble size of 2500
is over-cautious in this example and we could have used a smaller
ensemble size. We also believe that the required ensemble size to sample
from these posteriors is reduced by the use of the transport map.

The second obvious feature of these convergence plots is that the ETAIS
algorithms outperform the MH algorithms by a large margin - roughly a
reduction of two orders of magnitude in the relative error over the
same number of samples. A less positive result is that for this
example the ETAIS-TRW algorithm has found a map which has prevented
further convergence of the mean. This is likely because one dimension
of the transport map is sending proposals onto the negative real line
and so all samples are being assigned a zero weight. This is behaviour
which was anticipated, and it is always suggested that appropriate
preconditioners are used where the posterior is not absolutely
continuous with respect to Lebesgue measure, as is the case when the
posterior is only non-zero on the positive quadrant of $\mathbb{R}^d$
as in these examples. The large increase in the effective sample size
observed between the ETAIS-logRW and ETAIS-logTRW algorithms is
converted into an estimate which is twice as accurate after 1 million
samples.

@@

\section{Discussion}\label{sec:conc}
In this paper, we have investigated the use of transport maps to
accelerate and stabilise ensemble importance sampling schemes. In
particular we looked at the application of these methods to problems
arising in inverse problems for multiscale stochastic chemical
networks, where the posterior distributions which arise can be highly
concentrated along curved manifolds in parameter space.
Below we highlight some areas for future work.

@@

Algorithm~\ref{alg:TransportETAIS2} allows us to decorrelate the
dimensions of our random parameter on reference space, where we then
can resample and map the resulting ensemble back onto target
space. Since, on reference space, the dimensions are approximately uncorrelated, we
are able to resample in each dimension separately. Resampling in a
single dimension allows for optimisations in resampling code, and also
means that the resampler is not affected by the curse of
dimensionality.

In one
dimension the ETPF algorithm can be implemented very efficiently. As
described in~\cite{reich2013nonparametric}, the coupling matrix has
all non-zero entries in a staircase pattern when the state space is
ordered. We can exploit this knowledge to produce
Algorithm~\ref{alg:ETPF_1d}, which is much faster than using the
simplex algorithm to minimise the associated cost function, and faster
than the MT algorithm\cite{cotter2015parallel}. Moreover, the task of
running this resampler can then be parallelised over $d$
processors. This could be an area for further study.

@@

\subsection{Invertibility of the transport map}
In this paper, we have considered an approach similar to that
presented in \cite{parno2014transport}, where we aim to find an
invertible map which approximates the true transport map between the
reference measure and the target measure, however this true map is
invariably itself not invertible. However, by considering embedding
the map in higher dimensions\cite{whitney_self-intersections_1994,nash1956imbedding,takens_detecting_1981}, these issues of non-invertibility may be
avoided, and higher quality maps may be found which better aid
exploration and sampling of the whole state space.

>>>>

In this paper, we seek to develop an approach to
tackle this problem when using adaptive importance sampling methods,
by using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds. This approach is applicable to a whole range of
problems for which Monte Carlo Markov chain (MCMC) methods mix slowly.

@@

One common approach is to induce the quasi-equilibrium approximation
QEA. This approximation makes the assumption that fast reactions
enter quasi-equilibrium on a timescale which is negligible with
respect to the timescale on which the slow dynamics in the system are
evolving. This amounts to taking the asymptotic limit that the
propensities of the ``slow'' reaction channels are equal to zero. This
allows us to sample approximate trajectories of the slow reactions
without the need to compute many fast reaction events.

@@

We consider several different algorithms for sampling from these
distributions. First we implement both the ETAIS and MH algorithms with
a Gaussian proposal distribution. In the case of the ETAIS algorithm,
this is a Gaussian mixture proposal distribution, and in the MH
algorithm this is equivalent to a vanilla random walk. The proposal
distribution uses a covariance matrix which has been constructed using
the sample covariance of a sample produced using a standard MH-RW
algorithm. This is necessary since these algorithms otherwise fail to
converge in a reasonable time for
this problem. We also compare the ETAIS and MH algorithms when using a
transport map proposal distribution, as was discussed in detail in Chapter~\ref{sec:TETAIS}.

@@

The convergence analysis in the previous section was performed for the constrained
approximation of the posterior distribution. We now look at the
differences in the CMA and QEA approximations of the posterior distribution
arising in this problem. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
\[
	\hat{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{\alpha}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
\]
This difference in the denominator causes a shift in the manifold on
which the posterior is concentrated, as can be seen in
Figure~\ref{fig:chem_diff}, in which we plot the difference in the 
posteriors densities
\begin{equation}\label{eqn:chem_diff}\pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
\end{equation}

@@

Since the two posteriors have been approximated using an MCMC sample,
there is a visible amount of Monte Carlo static, particularly in the tails of
the distributions. We can see that the differences in the marginals
for $k_1, k_2,$ and $k_3$ are relatively small, but the marginal for
$k_4$ varies by a significant amount.

@@

Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals}
we see that we should of course prefer the model $\mathcal{M}_0$ in
which we observe all reactions perfectly. However this model is not
significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$,
~\cite{kass1995bayes}). The Bayes factor $B_{0,2} = 4.1 > 3.2$ tells
us that we should significantly prefer the full model to the QEA
approximation of the posterior. These Bayes factors provide evidence
for our
argument that the constrained approximation of the posterior is
superior to the QEA approximation.

@@

Figure~\ref{fig:chem_model_comp} displays more compelling evidence in
this direction. Here we show the marginal distributions for the
quantities $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$ with
respect to each of the three posterior distributions, $\pi_i(\cdot|D,
\mathcal{M}_i)$, $i = 0, 1, 2$. Firstly, in
Figure~\ref{fig:chem_model_comp} (a), displaying the marginals for
$\hat{k}_4^{\text{QEA}}$ we see that $\pi_0$, the density arising from
the full model with all reactions observed, is peaked a distance away
from $\pi_2$, QEA approximation. The CMA approximation $\pi_1$, in contrast, is
peaked close to peak of $\pi_0$, but there is more uncertainty here
since we do not observe the fast reactions.

In contrast, in Figure~\ref{fig:chem_model_comp} (b), the marginal
densities for $\hat{k}_4^{\text{CMA}}$ for $\pi_0$ and $\pi_1$ are
barely distinguishable, even with the vastly reduced data available in
the posterior $\pi_1$ with CMA approximation. The QEA approximation
$\pi_2$ is again peaked far away from both of the other distributions.
This demonstrates that the data regarding the slow variable
$S$ is only informative about $k_1 $ and $\hat{k}_4^{\text{CMA}}$, as
predicted by the constrained multiscale approximation.

\subsection{Gene Regulatory Network}\label{sec:grn}
In this example, we look at a model of a gene regulatory network (GRN)
mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering}
used by a cell to regulate the amount of a protein, $P$,
present. Proteins can bind in pairs to form a dimer $D$. The dimer can
bind to the gene $G$ and convert it into a ``switched off'' state $G'$, in which the
production of mRNA molecules $M$ is inhibited. The mRNA encodes for
production of the protein $P$, and both $P$ and $M$ can degrade. 
The eight reactions are given in Equation~\eqref{eqn:GRN_reactions}.
\begin{align}
	P+P \xleftrightarrows[k_2]{k_1} D \quad & \quad G+D \xleftrightarrows[k_4]{k_3} G' \nonumber\\
	G \xrightarrow{k_5} G+M \quad & \quad M\xrightarrow{k_6} M+P  \label{eqn:GRN_reactions} \\
	P \xrightarrow{k_7} \emptyset \quad & \quad M \xrightarrow{k_8} \emptyset \nonumber
\end{align}

We create trajectories of this system, using the Gillespie SSA, with the following parameters:
\begin{eqnarray*}
k_1 = 0.04, \qquad
k_2 = 5000, \qquad
k_3 = 100, \qquad
k_4 = 1, \\ 
k_5 = 0.5, \qquad
k_6 = 2, \qquad
k_7 = 0.2, \qquad
k_8 = 0.05,
\end{eqnarray*}
over the time period $t \in [0,500]$.

T-cells can be manufactured which bind to
proteins of a certain type, and cause them to phosphoresce, allowing
for approximate observations of the concentration levels of that
protein. However, the T-cells will not necessarily differentiate
between monomers and dimers of the protein. We approximately replicate such a
scenario, in a zero-observational noise setting, by assuming that we
can only observe $T=P + 2D + 2G'$, the total number of protein
molecules in the system, alongside $M$, the concentration of mRNA. The
number of dimers, and the status of the genetic switch, are assumed
unobservable. In
order to sample from the posterior distribution on the reaction
parameters in the system given these observations, we can either integrate over all possible
trajectories of the fast variables (which is computationally
intractable), or as in Section \ref{sec:chem_multiscale} we can use a
multiscale method to approximate the posterior.

@@

\subsubsection{Target Distribution}
As alluded to in the previous section, we aim to sample from the CMA
approximation of the posterior distribution as defined in
Equation~\eqref{eqn:chem_posterior}. This example is more involved
than the previous one, since we have an eight dimensional parameter
space corresponding to the eight reaction rates in the system
\eqref{eqn:GRN_reactions}. Half of those reactions are
unobservable, and their effect on the posterior is only felt through
their effect on the effective propensities $\hat{\alpha}_5$ and
$\hat{\alpha}_7$, and observations of the frequencies of these reactions.

The priors for this problem were chosen to be fairly uninformative,
and a list of the hyper-parameters corresponding to each Gamma prior can be found in Table~\ref{tab:grn_priors}.

@@

The one- and two-dimensional marginal distributions for the full
posterior distribution are displayed in
Figure~\ref{fig:GRN_posterior}. The posterior does not appear from
these joint marginals to contain many
interesting correlation structures, although several dimensions have
leptokurtic tails which are difficult for the standard ETAIS algorithm
to sample from. However as with the previous example, the density is
concentrated close to a lower-dimensional manifold. The marginal densities also vary over very different
scales, which could lead to problems with the algorithms which do not
benefit from the transport map.

@@

The transport map setup has not changed from the previous
example. However, due to the scaling of the weight we have had to be
more careful about which samples we use to build our map. We require
that the objective function $C(T)$, from
Equation~\eqref{eqn:TETAIS_objective} is convex, which requires that
the second derivative, given in Equation~\eqref{eqn:TETAIS_hessian},
is positive definite. This positive definite property depends on all
weights in our sample being strictly positive, which is not always
possible on a computer, where very small positive numbers can be
rounded down to absolute zero. For this reason we filter out any samples from the optimisation problem where the weight is a numeric zero. This does not affect the validity of the map since a weight zero would not contribute to the expectation.

@@

We believe that an ensemble size of 2500
is over-cautious, despite the larger number of dimensions in this
example, and that the required ensemble size to sample
from these posteriors is reduced by the use of the transport map.

The second obvious feature of these convergence plots is that the ETAIS
algorithms outperform the MH algorithms by a large margin - roughly a
reduction of two orders of magnitude in the relative error over the
same number of samples.

We note that in this 
example the ETAIS-TRW algorithm does not perform as well as the ETAIS-RW
algorithm. This is due to the lack of log preconditioner. The
posterior has support only on the positive quadrant, and without the
preconditioner, the transport map has to work hard to enforce
this. With a relatively low-dimensional approximation being used for
the transport map, this makes it more likely that the map produced
may hinder sampling rather than improving it. This highlights the
importance of using preconditioners with transport maps in order to
ensure absolute continuity of the posterior distribution with respect
to the pullback of the proposal distribution on reference space.

In contrast, when we look at the convergence of these two algorithms
with the log preconditioner, the transport map version exhibits more
consistent and stable convergence, with some of the 32 repeats for the
ETAIS-logRW algorithm producing rare large weights which hinder convergence.
The large increase in the effective sample size
observed between the ETAIS-logRW and ETAIS-logTRW algorithms is
converted into an estimate which is twice as accurate after 1 million
samples. This demonstrates our claim that the transport map can
produce stable ETAIS implementations with smaller ensemble sizes than without.

@@

\section{Discussion}\label{sec:conc}
In this paper, we have investigated the use of transport maps to
accelerate and stabilise ensemble importance sampling schemes. We
considered the ETAIS methodology here, but the approach used is
applicable to all such methods. The idea is powerful since it allows
adaptive importance sampling schemes to be efficient and stable for
smaller ensemble sizes. This in turn suggests that the improvements in
performance that adaptive importance sampling schemes can deliver
could be made applicable to a bigger class of problems.

In
particular we looked at the application of these methods to problems
arising in inverse problems for multiscale stochastic chemical
networks, where the posterior distributions which arise can be highly
concentrated along curved manifolds in parameter space. Through these
results we also demonstrated the improvement in accuracy of inference
for multiscale problems when more sophisticated multiscale
approximations are employed.
Below we highlight two areas for future work.

@@

Armed with a high quality transport map, we can transform the target
distribution close to a Gaussian reference measure, which is
uncorrelated in all dimensions. This lack of correlations allows us to
resample in each dimension separately. Resampling in a
single dimension allows for optimisations in resampling code, and also
means that the resampler is not affected by the curse of
dimensionality.

In one
dimension, the ETPF algorithm can be implemented very efficiently. As
described in~\cite{reich2013nonparametric}, the coupling matrix has
all non-zero entries in a staircase pattern when the state space is
ordered. We can exploit this knowledge to produce
Algorithm~\ref{alg:ETPF_1d}, which is much faster than using the
simplex algorithm to minimise the associated cost function, and faster
than the MT algorithm\cite{cotter2015parallel}, which is itself a
greedy approximation to the output of the ETPF resampler. Moreover, the task of
running this resampler can then be parallelised over $d$
processors. This could be an area for further study.

@@

\subsection{Invertibility of the transport map}
In this paper, we have considered an approach similar to that
presented in \cite{parno2014transport}, where we aim to find an
invertible map which approximates the true transport map between the
reference measure and the target measure. However, this true map is
invariably itself not invertible. However, by considering embedding
the map in higher dimensions\cite{whitney_self-intersections_1994,nash1956imbedding,takens_detecting_1981}, these issues of non-invertibility may be
avoided, and higher quality maps may be found which better aid
exploration and sampling of the whole state space.

\subsection{Adaptivity of ensemble sizes}
At initialisation, our transport map approximation is equal to the
identity, and the algorithm is essentially the standard ETAIS
algorithm. In order to build a good transport map approximation, we
require a good sample from the posterior distribution. Since our aim
is to apply this methodology to challenging problems, this is itself a
challenge. However, in the limit of the number of ensemble members
going to infinity, this approach is consistent and stable. However,
large ensemble sizes come with large computational costs.

We propose that the transport map acceleration could be used in
conjunction with an adaptive ensemble size mechanism. First we can
initialise the algorithm with a large ensemble size, and the identity transport map. As
we explore the density, the transport map quickly improves, making it
possible to sample efficiently and stably with a smaller ensemble
size. The problem then amounts to finding a reliable method to
approximate the lowest stable ensemble size.
